---
title: "Run PSA Simulations"
format: html
---

```{r}
set.seed(314)

library(tidyverse)
source("sim_functions.R")
source("polynomial_PSA.R")
```

## Where the paper's numbers come from

The paper's efficiency results (sample size for 80% power, the two-critique
decomposition, the regression-to-the-mean panel, and the adversarial
robustness check) come from a separate, faster engine, not from the
mechanism-based response-curve simulation in this document. That engine and its
two runners are:

- `lognormal_functions.R` — the clean log-normal generative model (PSA is
  log-normal; treatment shifts mean log follow-up) plus size-adjusted power.
- `run_efficiency.R` → `data/sim_results_efficiency.RDS` — the calibrated power
  curve, N* for 80% power, decomposition, and correlation/baseline-spread panel.
- `run_robustness.R` → `data/sim_results_robustness.RDS` — size-adjusted power
  under raw-additive effects and heavy-tailed error, the check that the
  log-scale advantage is not an artifact of the log-normal assumption.

Reproduce with `Rscript run_efficiency.R` and `Rscript run_robustness.R` from
this directory. We moved the primary comparison onto this model because the
response-curve mixture below is near-degenerate for a power comparison: its
deterministic curves make within-arm variance large relative to the between-arm
signal, so tests do not calibrate and the ranking reflects that rather than
information use. The response-curve simulation is retained for the
response-kinetics (AUC, half-life) work, which the log-normal single-timepoint
model does not address.

VARIABLES BEING CHANGED: month, correlation, responder curve distribution, and baseline PSA level

- months 3, 6, 9 and 12
- 5,000 times for null, 5,000 times for alternative
- 5,000 for null med, 5,000 for alt med
- 5,000 for null high, 5,000 for alt high
- correlations of 0.2, 0.4, 0.6, and 0.8
- month 3: 5,000 null w/ 1000 at 0.2 correlation, 1000 at 0.4, 1000 at 0.6, 1000 at 0.8. 5,000 alternative w/ same distribution
- same for months 6 9 and 12
- ultimately: 30,000 simulations for one month
-responder curves: e.g. grp A: 70% slow responder, 30% nonresponder, grp B: 40% slow responder, 60% nonresponder

TESTS
The two-critique framing (see decisions/2026-07-20-two-critique-reframe.md): PSA50 fails
both by dichotomizing (critique 1) and by using percent change as its scale (critique 2).
The tests below are ordered to isolate those effects. Fisher is the dichotomized status
quo; % change is the continuous-but-unadjusted, wrong-scale option; raw change and raw
ANCOVA move off percent change; the log-ratio t-test fixes the scale but is still an
unadjusted change score; log-scale ANCOVA fixes both (recommended); AUC/half-life capture
the response-kinetics information PSA50 discards regardless of scale.

fisher's exact: is there a difference in PSA50 between group A and group B?
% change: is the mean % change from baseline in grp A and grp B diferent? (ttest)
raw change: is the raw change from baseline in grp A and grp B different? (ttest)
ANCOVA (raw): after controlling for baseline, is there a difference in PSA change between A and B?
log-ratio: is the mean log ratio 100*(log(FUP)-log(baseline)) different between A and B? (ttest; unadjusted change score, kept as a comparator)
log-scale ANCOVA: after adjusting for log(baseline), is there a group difference in log(FUP)? (recommended)
AUC: is there a difference in RAW mean AUC between grp A and grp B? (ttest)
normalized AUC: is there a difference in adjusted AUC betwene grp A and grp B? (ttest)
half-life: no test, just present half lives for different scenarios

NULL HYPOTHESIS - NO DIFFERENCE BETWEN GRP A and B - RESPONDER CURVES
0.33, 0.33, 0.34 - 1
(treatment mildly effective)
0.60, 0, 0.40, - 2
0.40, 0, 0.60, -3 
(treatment highly effective)
0.80, 0, 0.20, - 4
0.20, 0, 0.80, - 5
(treatment unevenly effective)
0.5, 0.2, 0.3 - 6
0.2, 0.5, 0.3 - 7
0.2, 0.2, 0.6 - 8

ALTERNATIVE HYPOTHESIS - GRP A AND B DIFFER - RESPONDER CURVES
-note: showing that PSA50 is a metric that misses a lot of information
A: 0.33, 0.33, 0.35. B: 0.60, 0, 0.40
A: 0.40, 0, 0.60. B: 0.60, 0, 0.40
A: 0.20, 0, 0.80. B: 0.80, 0, 0.20
A: 0.2, 0.2, 0.6. B: 0.5, 0.2, 0.3
A: 0.2, 0.2, 0.6. B: 0.2, 0.5, 0.3
A: 0.33, 0.33, 0.34. B: 0.5, 0.2, 0.3
A: 0.33, 0.33, 0.35. B: 0.2, 0.5, 0.3
A: 0.2, 0.2, 0.6. B: 0.80, 0, 0.20
```{r}
# 1) 10, 10, 10
# 2) 18, 0, 12
# 3) 12, 0, 18
# 4) 24, 0, 6
# 5) 6, 0, 24
# 6) 15, 6, 9
# 7) 6, 15, 9
# 8) 6, 6, 18
props <- list(
  c(10, 10, 10), c(18, 0, 12), c(12, 0, 18), c(24, 0, 6),
  c(6, 0, 24), c(15, 6, 9), c(6, 15, 9), c(6, 6, 18)
)

props_sample <- seq(1:8)

# Each vector's components represent the type of response curve
# proportion for A and B respectively
props_alt <- list(
  c(1, 2), c(3, 2), c(5, 4), c(8, 6),
  c(8, 7), c(1, 6), c(1, 7), c(8, 4)
)

# left to right: 0.2, 0.4, 0.6, 0.8 correlation
low_corrs <- c(1.2, 0.55, 0.32, 0.18)
med_corrs <- c(7, 3.3, 2, 1.1)
high_corrs <- c(23.5, 11.1, 6.5, 3.7)

# defining results vectors for simulation
hypothesis <- rep(c(rep("Null", 15000), rep("Alternative", 15000)), 4)
a_prop <- c()
b_prop <- c()
baseline <- rep(rep(c("Low", "Medium", "High"), each = 5000), 8)
correlation <- rep(rep(c(0.2, 0.4, 0.6, 0.8), each = 1250), 24)
fisher_res <- c()
perchange_res <- c()
rawchange_res <- c()
ancova_res <- c()
log_ratio_res <- c()
log_ancova_res <- c()
raw_AUC_res <- c()
norm_AUC_res <- c()
half_life_res <- c()
```


## READ IN ANd TEST REAL DATA

```{r}
library(readxl)

psa_data <- read_excel("data/real_psa_data.xlsx",
                       skip = 1)

length(unique(psa_data$subject))

psa_data |> filter(subject == unique(subject)[35]) |>
  ggplot(aes(x = months, y = std_value)) + geom_point()
```

## Month 3

### Null Simulation

10,000 null w/ 2500 at 0.2 correlation, 2500 at 0.4, 2500 at 0.6, 2500 at 0.8. 10,000 alternative w/ same distribution
total for month 3 - 15,000 null, 15,000 alternative = 30,000
total for all sims at 3, 6, 9, and 12 - 120,000

- 5,000 times for null, 5,000 times for alternative
- 5,000 for null med, 5,000 for alt med
- 5,000 for null high, 5,000 for alt high
- correlations of 0.2, 0.4, 0.6, and 0.8
- month 3: 5,000 null w/ 1250 at 0.2 correlation, 1250 at 0.4, 1250 at 0.6, 1250 at 0.8. 5,000 alternative w/ same distribution
```{r}
############## Low Baseline PSA ##############
for (i in 1:5000) {
  # Correlation level: 1250 iterations at each of 0.2, 0.4, 0.6, 0.8
  corr_num <- ceiling(i / 1250)
  
  # Randomly samples a null response curve distribution for A and B
  response <- sample(props_sample, 1)
  a_prop <- c(a_prop, response)
  b_prop <- c(b_prop, response)
  
  # Dataframe of simulated PSA values for all months
  df_psa_sim <- generate_PSA_sim(
    4, 0.5, 4, 1, props[[response]], props[[response]], low_corrs[corr_num]
  )
  
  # Fisher's, %change, raw change, ANCOVA are calculated at individual timepoints
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  
  # raw AUC, normalized AUC, and half-life are calculated as a whole
  # each is calculated individually and then averaged
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  # add simulated test results to overall results
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}

############## Medium Baseline PSA ##############
for (i in 1:5000) {
  corr_num <- ceiling(i / 1250)
  
  response <- sample(props_sample, 1)
  a_prop <- c(a_prop, response)
  b_prop <- c(b_prop, response)
  
  df_psa_sim <- generate_PSA_sim(
    20, 3, 20, 3, props[[response]], props[[response]], med_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
  
}

############## High Baseline PSA ##############
for (i in 1:5000) {
  corr_num <- ceiling(i / 1250)
  
  response <- sample(props_sample, 1)
  a_prop <- c(a_prop, response)
  b_prop <- c(b_prop, response)
  
  df_psa_sim <- generate_PSA_sim(
    100, 10, 100, 10, props[[response]], props[[response]], high_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  

  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
  
}
```



### Alternative Simulation

```{r}
############## Low Baseline PSA ##############
for (i in 1:5000) {
  # Correlation level: 1250 iterations at each of 0.2, 0.4, 0.6, 0.8
  corr_num <- ceiling(i / 1250)
  
  # Randomly samples an alternative 
  # response curve distribution for A and B
  response <- sample(props_alt, 1)[[1]]
  a_prop <- c(a_prop, response[1])
  b_prop <- c(b_prop, response[2])
  
  df_psa_sim <- generate_PSA_sim(
    4, 0.5, 4, 1, props[[response[1]]], props[[response[2]]], low_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}


############## Medium Baseline PSA ##############
for (i in 1:5000) {
  corr_num <- ceiling(i / 1250)
  
  response <- sample(props_alt, 1)[[1]]
  a_prop <- c(a_prop, response[1])
  b_prop <- c(b_prop, response[2])
  
  df_psa_sim <- generate_PSA_sim(
    20, 3, 20, 3, props[[response[1]]], props[[response[2]]], med_corrs[corr_num]
  )

  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}


############## High Baseline PSA ##############
for (i in 1:5000) {
  corr_num <- ceiling(i / 1250)
  
  response <- sample(props_alt, 1)[[1]]
  a_prop <- c(a_prop, response[1])
  b_prop <- c(b_prop, response[2])
  
  df_psa_sim <- generate_PSA_sim(
    100, 10, 100, 10, props[[response[1]]], props[[response[2]]], high_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}
```

### Summarize and save month 3 results

Rejection rate at the trial's nominal Type I level (0.10): under the null this is the
realized Type I error, under the alternative it is power. Saved for the paper to read
(`index.qmd` loads this file rather than re-running the simulation).

```{r}
alpha <- 0.10
n_done <- length(fisher_res)          # 30000 once month 3's null + alt loops finish
idx <- seq_len(n_done)

df_results_month3 <- tibble(
  hypothesis  = hypothesis[idx],
  baseline    = factor(baseline[idx], levels = c("Low", "Medium", "High")),
  correlation = correlation[idx],
  fisher      = fisher_res,
  perchange   = perchange_res,
  rawchange   = rawchange_res,
  ancova      = ancova_res,
  log_ratio   = log_ratio_res,
  log_ancova  = log_ancova_res
)

gts_power_month3 <- df_results_month3 |>
  pivot_longer(fisher:log_ancova, names_to = "test", values_to = "p") |>
  summarise(
    .by = c(hypothesis, baseline, correlation, test),
    reject_rate = mean(p < alpha, na.rm = TRUE),
    n_sims      = sum(!is.na(p))
  )

df_kinetics_month3 <- tibble(
  hypothesis = hypothesis[idx],
  baseline   = factor(baseline[idx], levels = c("Low", "Medium", "High")),
  raw_auc    = raw_AUC_res,
  norm_auc   = norm_AUC_res,
  half_life  = half_life_res
) |>
  summarise(
    .by = c(hypothesis, baseline),
    raw_auc   = mean(raw_auc, na.rm = TRUE),
    norm_auc  = mean(norm_auc, na.rm = TRUE),
    half_life = mean(half_life, na.rm = TRUE)
  )

saveRDS(
  list(power = gts_power_month3, kinetics = df_kinetics_month3, alpha = alpha),
  "data/sim_results_month3.RDS"
)
```



## Month 6



## Month 9



## Month 12