---
title: "Run PSA Simulations"
format: html
---

```{r}
set.seed(314)

library(tidyverse)
source("sim_functions.R")
source("polynomial_PSA.R")
```
VARIABLES BEING CHANGED: month, correlation, responder curve distribution, and baseline PSA level

- months 3, 6, 9 and 12
- 5,000 times for null, 5,000 times for alternative
- 5,000 for null med, 5,000 for alt med
- 5,000 for null high, 5,000 for alt high
- correlations of 0.2, 0.4, 0.6, and 0.8
- month 3: 5,000 null w/ 1000 at 0.2 correlation, 1000 at 0.4, 1000 at 0.6, 1000 at 0.8. 5,000 alternative w/ same distribution
- same for months 6 9 and 12
- ultimately: 30,000 simulations for one month
-responder curves: e.g. grp A: 70% slow responder, 30% nonresponder, grp B: 40% slow responder, 60% nonresponder

TESTS
fisher's exact: is there a difference in PSA50 between group A and group B?
% change: is the mean % change from baseline in grp A and grp B diferent? (ttest)
raw change: is the raw change from baseline in grp A and grp B different? (ttest)
ANCOVA: after controlling for baseline, is there a significant difference in the change in PSA between grp A and grp B?
AUC: is there a difference in RAW mean AUC between grp A and grp B? (ttest)
normalized AUC: is there a difference in adjusted AUC betwene grp A and grp B? (ttest)
half-life: no test, just present half lives for different scenarios

NULL HYPOTHESIS - NO DIFFERENCE BETWEN GRP A and B - RESPONDER CURVES
0.33, 0.33, 0.34 - 1
(treatment mildly effective)
0.60, 0, 0.40, - 2
0.40, 0, 0.60, -3 
(treatment highly effective)
0.80, 0, 0.20, - 4
0.20, 0, 0.80, - 5
(treatment unevenly effective)
0.5, 0.2, 0.3 - 6
0.2, 0.5, 0.3 - 7
0.2, 0.2, 0.6 - 8

ALTERNATIVE HYPOTHESIS - GRP A AND B DIFFER - RESPONDER CURVES
-note: showing that PSA50 is a metric that misses a lot of information
A: 0.33, 0.33, 0.35. B: 0.60, 0, 0.40
A: 0.40, 0, 0.60. B: 0.60, 0, 0.40
A: 0.20, 0, 0.80. B: 0.80, 0, 0.20
A: 0.2, 0.2, 0.6. B: 0.5, 0.2, 0.3
A: 0.2, 0.2, 0.6. B: 0.2, 0.5, 0.3
A: 0.33, 0.33, 0.34. B: 0.5, 0.2, 0.3
A: 0.33, 0.33, 0.35. B: 0.2, 0.5, 0.3
A: 0.2, 0.2, 0.6. B: 0.80, 0, 0.20
```{r}
# 1) 10, 10, 10
# 2) 18, 0, 12
# 3) 12, 0, 18
# 4) 24, 0, 6
# 5) 6, 0, 24
# 6) 15, 6, 9
# 7) 6, 15, 9
# 8) 6, 6, 18
props <- list(
  c(10, 10, 10), c(18, 0, 12), c(12, 0, 18), c(24, 0, 6),
  c(6, 0, 24), c(15, 6, 9), c(6, 15, 9), c(6, 6, 18)
)

props_sample <- seq(1:8)

# Each vector's components represent the type of response curve
# proportion for A and B respectively
props_alt <- list(
  c(1, 2), c(3, 2), c(5, 4), c(8, 6),
  c(8, 7), c(1, 6), c(1, 7), c(8, 4)
)

# left to right: 0.2, 0.4, 0.6, 0.8 correlation
low_corrs <- c(1.2, 0.55, 0.32, 0.18)
med_corrs <- c(7, 3.3, 2, 1.1)
high_corrs <- c(23.5, 11.1, 6.5, 3.7)

# defining results vectors for simulation
hypothesis <- rep(c(rep("Null", 15000), rep("Alternative", 15000)), 4)
a_prop <- c()
b_prop <- c()
baseline <- rep(rep(c("Low", "Medium", "High"), each = 5000), 8)
correlation <- rep(rep(c(0.2, 0.4, 0.6, 0.8), each = 1250), 24)
fisher_res <- c()
perchange_res <- c()
rawchange_res <- c()
ancova_res <- c()
log_ratio_res <- c()
log_ancova_res <- c()
raw_AUC_res <- c()
norm_AUC_res <- c()
half_life_res <- c()
```


## READ IN ANd TEST REAL DATA

```{r}
library(readxl)

psa_data <- read_excel("data/real_psa_data.xlsx",
                       skip = 1)

length(unique(psa_data$subject))

psa_data |> filter(subject == unique(subject)[35]) |>
  ggplot(aes(x = months, y = std_value)) + geom_point()
```

## Month 3

### Null Simulation

10,000 null w/ 2500 at 0.2 correlation, 2500 at 0.4, 2500 at 0.6, 2500 at 0.8. 10,000 alternative w/ same distribution
total for month 3 - 15,000 null, 15,000 alternative = 30,000
total for all sims at 3, 6, 9, and 12 - 120,000

- 5,000 times for null, 5,000 times for alternative
- 5,000 for null med, 5,000 for alt med
- 5,000 for null high, 5,000 for alt high
- correlations of 0.2, 0.4, 0.6, and 0.8
- month 3: 5,000 null w/ 1250 at 0.2 correlation, 1250 at 0.4, 1250 at 0.6, 1250 at 0.8. 5,000 alternative w/ same distribution
```{r}
############## Low Baseline PSA ##############
corr_num <- 1

for (i in 1:5000) {
  # Changes correlation to simulate at
  if (i %% 1250 == 0) {
    corr_num <- corr_num + 1
  }
  
  # Randomly samples a null response curve distribution for A and B
  response <- sample(props_sample, 1)
  a_prop <- c(a_prop, response)
  b_prop <- c(b_prop, response)
  
  # Dataframe of simulated PSA values for all months
  df_psa_sim <- generate_PSA_sim(
    4, 0.5, 4, 1, props[[response]], props[[response]], low_corrs[corr_num]
  )
  
  # Fisher's, %change, raw change, ANCOVA are calculated at individual timepoints
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  
  # raw AUC, normalized AUC, and half-life are calculated as a whole
  # each is calculated individually and then averaged
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  # add simulated test results to overall results
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}

############## Medium Baseline PSA ##############
corr_num <- 1

for (i in 1:5000) {
  if (i %% 1250 == 0) {
    corr_num <- corr_num + 1
  }
  
  response <- sample(props_sample, 1)
  a_prop <- c(a_prop, response)
  b_prop <- c(b_prop, response)
  
  df_psa_sim <- generate_PSA_sim(
    20, 3, 20, 3, props[[response]], props[[response]], med_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
  
}

############## High Baseline PSA ##############
corr_num <- 1

for (i in 1:5000) {
  if (i %% 1250 == 0) {
    corr_num <- corr_num + 1
  }
  
  response <- sample(props_sample, 1)
  a_prop <- c(a_prop, response)
  b_prop <- c(b_prop, response)
  
  df_psa_sim <- generate_PSA_sim(
    100, 10, 100, 10, props[[response]], props[[response]], high_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  

  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
  
}
```



### Alternative Simulation

```{r}
############## Low Baseline PSA ##############
corr_num <- 1

for (i in 1:5000) {
  # Changes correlation to simulate at
  if (i %% 1250 == 0) {
    corr_num <- corr_num + 1
  }
  
  # Randomly samples an alternative 
  # response curve distribution for A and B
  response <- sample(props_alt, 1)[[1]]
  a_prop <- c(a_prop, response[1])
  b_prop <- c(b_prop, response[2])
  
  df_psa_sim <- generate_PSA_sim(
    4, 0.5, 4, 1, props[[response[1]]], props[[response[2]]], low_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}


############## Medium Baseline PSA ##############
corr_num <- 1

for (i in 1:5000) {
  if (i %% 1250 == 0) {
    corr_num <- corr_num + 1
  }
  
  response <- sample(props_alt, 1)[[1]]
  a_prop <- c(a_prop, response[1])
  b_prop <- c(b_prop, response[2])
  
  df_psa_sim <- generate_PSA_sim(
    20, 3, 20, 3, props[[response[1]]], props[[response[2]]], med_corrs[corr_num]
  )

  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}


############## High Baseline PSA ##############
corr_num <- 1

for (i in 1:5000) {
  if (i %% 1250 == 0) {
    corr_num <- corr_num + 1
  }
  
  response <- sample(props_alt, 1)[[1]]
  a_prop <- c(a_prop, response[1])
  b_prop <- c(b_prop, response[2])
  
  df_psa_sim <- generate_PSA_sim(
    100, 10, 100, 10, props[[response[1]]], props[[response[2]]], high_corrs[corr_num]
  )
  
  df_psa_wide <- df_psa_sim |> filter(time == 0 | time == 3) |>
      pivot_wider(names_from = time, values_from = PSA) |>
      mutate(baseline = `0`,
             FUP = `3`,
             change = baseline - FUP,
             psa50_achieved = (change/baseline) >= 0.50
      )
  
  fisher_sim <- fisher_sim_test(df_psa_wide)
  perchange_sim <- perchange_sim_test(df_psa_wide)
  rawchange_sim <- raw_change_test(df_psa_wide)
  ancova_sim <- ancova_test(df_psa_wide)
  log_ratio_sim <- log_percent_test(df_psa_wide)
  log_ancova_sim <- log_ancova_test(df_psa_wide)
  raw_auc <- raw_auc_calc(df_psa_sim)
  norm_auc <- norm_auc_calc(df_psa_sim)
  halflife <- halflife_calc(df_psa_sim)
  
  fisher_res <- c(fisher_res, fisher_sim)
  perchange_res <- c(perchange_res, perchange_sim)
  rawchange_res <- c(rawchange_res, rawchange_sim)
  ancova_res <- c(ancova_res, ancova_sim)
  log_ratio_res <- c(log_ratio_res, log_ratio_sim)
  log_ancova_res <- c(log_ancova_res, log_ancova_sim)
  raw_AUC_res <- c(raw_AUC_res, raw_auc)
  norm_AUC_res <- c(norm_AUC_res, norm_auc)
  half_life_res <- c(half_life_res, halflife)
}
```



## Month 6



## Month 9



## Month 12